\stdpopsim: a community toolkit for population genetic simulation

### 2019-12-11 popSimManu.tex

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
As large databases of population genetic variation begin to be used to inform public health procedures,
the accuracy and quality of these inferences is becoming even more important.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \texttt{stdpopsim}, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

Benchmarking the performance of inference methods in population genetics is not straightforward.
Different methods often fit different types of model
or use different summaries of the genetic variation data, even when
attempting to address the same biological question. For example, there are
numerous methods to infer population size changes.
The \MSMC method (\cite{schiffels2014inferring}) allows for the
population to continuously change in size over time. Other methods, such as
\dadi (\cite{gutenkunst2009inferring}), allow for a smaller number of size changes to occur at certain
points throughout the population's history. The types of data that the
methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
spatial patterns of genetic variation along the sequence, \dadi  instead uses the
site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
sample, ignoring any information about the correlation structure (linkage
disequilibrium) among SNPs. Thus from the start these methods may not be
expected to perform equally well in all scenarios.

The degree to which modeling assumptions and different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
suggested large ancient ($>100$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see
\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

One way to benchmark different methods for statistical inference in population
genetics is to apply the methods to simulated datasets where the true model
parameters are known. Methods that perform well will accurately infer these
parameters and will also provide reasonable estimates of uncertainty of the
estimated parameters. Indeed, simulations studies are a standard ingredient in
most papers proposing new methods in population genetics. However, these
simulation studies often use simulations of the particular models for which the method was designed,
which showcases the novel method rather than rigorously comparing performance to competing methods.
Of course, testing a new method in the situation for which it was designed is an important step,
but more rigorous testing is necessary for tools that are to be widely used in the field.
% Moreover researchers are generally forced to reimplement previously
% inferred scenarios independently, which makes good benchmarking
% a major additional time investment, and can lead to errors.
% Lastly, the degree to which different methods are compared can be quite variable.

One reason why simulation studies do not compare the performance of many methods
across a wide range of situations
is that simulation studies are challenging to carry out, for several reasons.
First, while there are many published population genetic models available, their
parameters are often not easily accessible, usually presented in supplemental
tables. Translating these parameters into input for a given simulation engine can
be a time-consuming and error prone process. In addition, papers often present many
different models, and different authors may choose to select distinct models,
making comparison difficult. While improvements in simulation software and ever
faster computers have partially ameliorated this concern, generating the
simulated datasets can still be a time and computing resource bottleneck.

@@

The software has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and accuracy. Below we describe the resource and give an
example of how it can be used to benchmark demographic inference methods.

@@

>>>>

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
As large databases of population genetic variation begin to be used to inform public health procedures,
the accuracy and quality of these inferences is becoming even more important.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \stdpopsim, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

The degree to which modeling assumptions and different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
suggested large ancient ($>100$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see
\citet{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

One way to benchmark different methods for statistical inference in population
genetics is to apply the methods to simulated datasets where the true model
parameters are known. Methods that perform well will accurately infer these
parameters and will also provide reasonable estimates of uncertainty of the
estimated parameters. Indeed, simulations studies are a standard ingredient in
most papers proposing new methods in population genetics. However, these
simulation studies often use simulations of the particular models for which the method was designed,
which showcases the novel method rather than rigorously comparing performance to competing methods.
Of course, testing a new method in the situation for which it was designed is an important step,
but more rigorous testing is necessary for tools that are to be widely used in the field.
% Moreover researchers are generally forced to reimplement previously
% inferred scenarios independently, which makes good benchmarking
% a major additional time investment, and can lead to errors.
% Lastly, the degree to which different methods are compared can be quite variable.
Many more researchers in the broader field
use simulations to perform power and other exploratory analyses.
The many steps of implementing a moderately complicated population genetics simulation,
including translating published demographic models to a simulator
and obtaining genetic maps,
present a major obstacle, especially to new researchers.

% One reason why simulation studies do not compare the performance of many methods
% across a wide range of situations
% is that simulation studies are challenging to carry out, for several reasons.
% First, while there are many published population genetic models available, their
% parameters are often not easily accessible, usually presented in supplemental
% tables. Translating these parameters into input for a given simulation engine can
% be a time-consuming and error prone process. In addition, papers often present many
% different models, and different authors may choose to select distinct models,
% making comparison difficult. While improvements in simulation software and ever
% faster computers have partially ameliorated this concern, generating the
% simulated datasets can still be a time and computing resource bottleneck.

@@

The software has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and accuracy. 
Importantly, we have rigorous quality control methods to ensure implemented models are accurate,
and have documented methods for others to contribute new modules.
We welcome new collaborators.
Below we describe the resource and give an
example of how it can be used to benchmark demographic inference methods.

@@

\paragraph{The problem of comparison} % Moved from the intro
Benchmarking the performance of inference methods in population genetics is not straightforward.
Different methods often fit different types of model
or use different summaries of the genetic variation data, even when
attempting to address the same biological question. For example, there are
numerous methods to infer population size changes.
The \MSMC method \citep{schiffels2014inferring} allows for the
population to continuously change in size over time. Other methods, such as
\dadi \citep{gutenkunst2009inferring}, allow for a smaller number of size changes to occur at certain
points throughout the population's history. The types of data that the
methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
spatial patterns of genetic variation along the sequence, \dadi  instead uses the
site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
sample, ignoring any information about the correlation structure (linkage
disequilibrium) among SNPs. Thus from the start these methods may not be
expected to perform equally well in all scenarios,
and indeed, it is not always clear how to compare to the truth --
for instance, how well can a method of piecewise constant population sizes
describe a period of exponential growth?
Or, how should discrete population models be mapped onto real-world (continuous) geography?
The answer depends on determining what is most important for the question at hand.
